### 2019-11-09 paper/conceptual_framework.tex
Expand on mortality risk on spacing and clean up of notes

Biological factors are, however, unlikely to be an important factor in differences in spacing over 
time or across education levels in India for the period covered.
Outside of direct famine conditions, there is little evidence that women's health and 
nutritional status substantially affects spacing \citep{Huffman1987,John1987,lindstrom99}.
The Maharashtra drought occurred in 1970-1973 and, therefore, would only affect the first
year or two of the data.
Hence, although there are still differences in individual fecundity across women, it is
unlikely that we should see systematic differences across education groups and over time
in the data because of health status.
Furthermore, I only examine spacing up to parity fourth, and age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, when though better 
educated women are, on average, older when they marry.

@@

Most of the research finds a negative impact of short spacing on child health, especially 
for very short intervals of 24 months or less 
\citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
The effects of short spacing on child mortality are, however, strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Molitoris2019}.

@@

Education affects a number of factors related to birth spacing.
First, it affects the cost of time and the associated income profile over time.
Second, it affects the cost of using contraceptives (independently of the lower fertility
and changes in spacing). This could also include how good women are to use new 
contraceptive technology \citep{Kim2010}.
[I find this argument weird. \citep{Rosenzweig1989} shows that better educated women
are better at using more difficult ``technologies'' but those do not include most common
forms of modern contraceptives such as pills or injectables.]

First, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Second, if there is a risk of ``overshooting'' the ideal number of children because of
imperfect fertility control and overshooting is sufficiently costly.
Parents would select a higher level of contraception than if they had perfect control and,
consequently, space children far apart on average [this is presumably
the \citep{Keyfitz1971,Heckman1976} arguments]
Third, if capital markets are not perfect and giving birth or rearing a child is costly
parents are forced to wait until they have enough income/wealth to have the next child
(this works through consumption; with low levels of consumption the cost of giving up
any will be higher so parents will want to wait until they have higher income and the
marginal cost of giving up consumption will be lower).
[would the opportunity cost of time work in the opposite direction? If women's 
opportunity cost of time goes up over time it would make sense to have the children
earlier]

Is there an expected mortality effect here as well?
If you expect a certain proportion of your children to die, it would probably make sense
to have your births earlier, everything else equal.
That way, you would get to the ideal number of children earlier.

@@

Biological/health effects on spacing that might change with education (or rather the
health improvements that come with higher education/income either from the women
herself or her husband)?
As better health leads to shorter post-partum amenorrhoea, we should expect shorter
spacing rather than longer with better health.
The evidence suggests that there is relatively little effect of health status on 
fecundity once nutrition is above famine or starvation levels \citep{Huffman1987,John1987,lindstrom99}

Breastfeeding would be part of this.
An healthier mother is more likely to successfully breastfeed and that would increase the
spacing between births.

>>>>

Biological factors are, however, unlikely to be an important factor in differences in spacing over 
time or across education levels in India for the period covered.
Outside of direct famine conditions, there is little evidence that women's health and 
nutritional status substantially affects spacing \citep{Huffman1987,John1987,lindstrom99}.%
\footnote{
On one hand, with better health we should expect shorter post-partum amenorrhoea and a higher
likelihood of carrying a pregnancy to term, both of which lead to shorter spacing with better health.
On the other hand, an healthier mother is more likely to successfully breastfeed and that would 
increase the spacing between births, although if she is a strong producer of breastmilk the 
suckling relative to production is lower, which, in turn, lower the conceptraceptive effect
of breastfeeding.
}
The Maharashtra drought occurred in 1970-1973 and, therefore, would only affect the first
year or two of the data.
Hence, although there are still differences in individual fecundity across women, it is
unlikely that we should see systematic differences across education groups and over time
in the data because of health status.
Furthermore, I only examine spacing up to parity fourth, and age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, when though better 
educated women are, on average, older when they marry.

@@

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Molitoris2019}.

@@

@@
